---
title: "Problem Set 03: Data Wrangling"
author: "Math 271"
date: "Spring 2022"
output:
  html_document:
    theme: lumen
    toc: yes
    toc_float: yes
    df_print: paged
    css: lab.css
---

In this problem set we will practice some of the key data manipulation tasks for describing, summarizing, and working with data. This material is covered at greater length in [Chapter 3: Data Wrangling](https://moderndive.com/3-wrangling.html) We will specifically review the following functions from the `dplyr` package: 

-  `select`
-  `mutate`
-  `summarize`
-  `arrange`
-  `filter`
-  `group_by`
-  `%>%`

In addition we will review how to save objects using the `<-` assignment operator. 

# Set up

As before, you should use the Laulima link to accept the assignment on Github, then create a new project in RStudio that clones your new Github repository. See the first problem set for detailed instructions. 

The solution file for this assignment should be named `PS03_solution.Rmd`.

### R Packages
Run the following to load the necessary packages for this problem set:
```{r warning = F, message = F}
library(ggplot2)
library(dplyr)
```

### The data
Run the following to load and take a `glimpse` of the data.
```{r eval = F}
data(txhousing)
glimpse(txhousing)
```

This is an example data set that is built into the `ggplot2` package. It is about housing in Texas. Each row is monthly data for a given city in Texas in a given year. There are multiple years of data for each city. The `data` command loads built-in data sets into your working environment.

Most built-in data sets have a help file that describes the data. Take a look at the the help file for the data by typing 
```{r eval=F}
?txhousing
``` 
into the **console**. What is the `listings` variable in this data set?

***
# DATA WRANGLING REVIEW

## pipe (`%>%`)

The pipe is an _infix_ operator, like `+`. It takes a left-hand-side (lhs) and a right-hand-side (rhs) and does something with them, producing a result. The pipe does something quite unusual though: it converts `x %>% f` into `f(x)`, returning the result of the function call. The benefit is not in a single function call, but in the way that commands can be _composed_ or _piped_ together. For example, the command
```{r eval=F}
x %>% f %>% g %>% h
```
means, in prose, to take object `x` and apply the function `f` to it, take the output from that and apply function `h` to it, then take the output from that and apply the function `h` to it, finally returning the result.
It is equivalent to 
```{r eval=F}
h(g(f(x)))
```
which is more difficult to read and understand logically,
and also to 
```{r eval=F}
temp1 <- f(x)
temp2 <- g(temp1)
h(temp2)
```
which is somewhat wasteful, difficult to maintain, and hard to insert or remove steps.

Even more usefully, if the right side is a full function call, it will insert the left hand side as the first argument before running it `x %>% f(y, z)` is equivalent to `f(x, y, z)`. A single `.` can be used as a placeholder if you need the left side to go somewhere other than the first argument. So `x %>% f(y, ., z)` is equivalent to `f(y, x, z)`

**The pipe operator is a bit of a pain to type, so in RStudio, it can be inserted with hotkey `Ctrl-Shift-M`.**

## select
Sometimes we want to pull out or extract just one or two **columns** of data. The following code will extract only the column in the data set for the variables `sales` and `volume`. Optionally, it can rename a variable while selecting it, shortening `volume` to `vol`.
```{r eval = F}
txhousing %>% select(sales, vol = volume)
```

Note that we did not use an assignment operator `<-` so we did not save these extracted, selected values. In the following code, we **save** the results, in a data frame **ASLO** called txhousing. By putting `-` in front of the `date` variable we tell R to select all **except** the `date` variable. 
Run the following code:

```{r eval=F}
txhousing <- txhousing %>% select(-date)
```

If you look at txhousing in the data viewer, the `date` variable is no longer included. 

To restore the original `txhousing` data with all columns, use the "brush" button in the Environment pane to remove your modified variable, or use the command `rm(txhousing)`.

Have a look at the help `?select` for some hints on other ways to use the command. Some highlights:

- `select(city:volume)` select all columns starting at `city` and ending at `volume`
- `starts_with("string")` get all columns that start with a given string. There are several similar functions for searching in different ways.

The examples at the end of the help file are very useful.

## filter

The filter function allows you to pull out just the **rows** (cases or observations) you want, based on some criteria in a column value.

Imagine for instance that we wanted to reduce the data set include data for only 2012, in Austin. This code chunk takes the `txhousing` data, then filters it to only include rows in which the year is 2012, **and** the city is Austin. The results are saved in a new data frame called `austin_12`. Variable assignment is usually silent, but sometimes you do want to view the thing that is saved. A useful trick to achieve this is to wrap the entire assignment command in a pair of parentheses.

```{r }
( austin_12 <- txhousing %>% filter(year == 2012, city == "Austin") )
```
 
> Note that we use `==` to identify the desired criteria. Try changing one of the `==` to a single `=` and see what happens when you run it.

What if we wanted to restrict our data set to only years before 2004 and the City of Austin? Below we use the `<` symbol to accomplish this. Note we did not **SAVE** these results in a new data frame...so no new data frame showed up in our Environment pane, but the results print out immediately below the code chunk. 

```{r eval = F}
txhousing %>% filter(year < 2004, city == "Austin")
```

What if we wanted to use multiple cities?  Below we use the  `|` symbol to indicate that the city could be Austin **OR** Abilene. In this case, we **saved** these results as a new data frame called `aust_ab` that appears in your Environment pane.  

```{r }
aust_ab <- txhousing %>% filter(city == "Austin" | city == "Abilene")
```

Have a look at the help `?filter` for more info. The help page will likely tell you there are multiple commands named `filter`. You are interested in _Subset rows using column values_ in the _dplyr_ package.

## mutate
The mutate function can add new columns (variables) to a data frame or alter existing ones. For instance, the following will add a new column to the data called `vol_100k` that expresses volume in units of $100000. 
```{r}
txhousing <- txhousing %>%
  mutate(vol_100k = volume/100000)
```

Note that we **SAVED** these results in new data frame called `txhousing`. This therefore **overwrote** the old `txhousing` data frame with a new version that contains this column. You can open the `txhousing` data frame in the viewer to confirm that it now contains this new column. 

## summarize
One of the first tasks in data analysis is often to get descriptive statistics that help to understand the central tendency and variability in the data. The `summarize()` command can take a column of data, and reduce it to a summary statistic. 

For instance, the code below uses the `austin_12` data set made earlier to calculate the mean monthly number of `sales` in Austin in 2012. 

```{r eval = F}
austin_12 %>% summarize(x_bar_sales = mean(sales))
```
This code tells R to calculate the `mean` of the variable `sales`, and to save the results in a variable called `x_bar_sales`. 

You can also calculate multiple summary statistics at once, and even for multiple variables. Below we also calculate a standard deviation `sd()` of `sales`, a minimum `min()` of the `volume` variable, a maximum `max()` of the `volume` variable, etc. The `n()` calculates sample size...or the number of rows/ cases in the data frame. 

```{r }
austin_12 %>% summarize(x_bar_sales = mean(sales), 
                        sd_sales = sd(sales), 
                        min_vol = min(volume), 
                        max_vol = max(volume), 
                        mdn_list = median(listings), 
                        iqr_list = IQR(listings),
                        sample_size = n())
```

Note that the names of the elements you calculate are  user defined, like `xbar_sales`, `min_vol`, and `mdn_list`. You could customize these names as you like (but don't use spaces in your names). 

## arrange
You just determined that the maximum volume of monthly sales in Austin in 2012 was a total of $791,281,075 ....but what if you wanted to know **WHAT MONTH** that occurred in? Copy paste, and run the following into a new code chunk: 

```{r eval = F}
austin_12 %>%
  arrange(desc(volume))
```

This tells R to `arrange` the rows in the data set based on the `volume` column, and to do so in `desc`ending order. So the row with the $791,281,075 in sales is shown at the top! We can see that this `volume` occurred in the sixth month (June). 

## group_by
Sometimes we also want to calculate summary statistics across different levels of another variable. For instance, here we find the average number of monthly sales that occurred in Abilene and Austin across all years in the data set. Note that we **use the `aust_ab` data frame** we created earlier, to restrict our analysis to those two cities. 

```{r}
aust_ab %>% group_by(city) %>% 
  summarize(x_bar_sales = mean(sales))
```

From the results we can see that there were an average of 150 sales per month in Abilene, and 1996 in Austin. 

We can give R multiple variables to group by. For instance, this code gives us the mean sales for each month in each city, averaged across all the years. So for instance the mean number of sales in January, in Abilene was 96 homes. 
```{r eval = F}
aust_ab %>% group_by(city, month) %>% 
  summarize(x_bar_sales = mean(sales))
```

## `slice` and friends

The `slice` command extracts rows by number (index). So `txhousing %>% slice(5)` gives you the 5th row of the data. The general form is not quite as useful as the friends `slice_head` and `slice_tail`, which take the first/last rows of the data. 


```{r eval=F}
txhousing %>% slice_head(n=5) # first 5 rows
txhousing %>% slice_tail(prop=0.001) # the final 0.1% of the data
```

Slice can be useful for data exploration, especially after `arrange`. If the data is grouped, the slice operates within the groups.

This filters to years before 2002, then groups by city and year, and returns only the second and third row in each group.
```{r}
txhousing %>% filter(year<2002) %>% group_by(city,year) %>% slice(2:3)
```


***
# Independent practice 

## Basic syntax
This first set of questions will help you practice basic syntax. All you need to include is a Question header, and code for each. 

4. Make a data set called `dallas_sub` that includes data only from the city of Dallas in 2012 & 2013. 


5.
Add a column **to the `dallas_sub` data set** called `prct_sold` that calculates the percentage of `listings` that were `sold` (`sales/listings * 100`). Be sure to **save** the results also as a data frame called `dallas_sub`. 


6.
Calculate the **average** percentage of listings that were sold in Dallas **in each month of the year** (`group_by`) based on your `dallas_sub` data set. Save the results of the calculation in an data frame called `dallas_summary`. 


7.
Arrange the `dallas_summary` in `desc`ending order based on the average percentage of listings that were sold in Dallas in 2012, so you can see **which month** had the greatest percentage of houses sold in Dallas on average from 2012-2013.  You do not need to save the results. 


***
## More advanced wrangling


8.
Run the following code chunk. Study the code, and the output. Explain in your own words what this code chunk calculated. 
```{r eval = F}
txhousing %>% 
  filter(year == 2012 | year == 2013, city == "Dallas") %>%
  mutate(prct_sold = sales/listings *100) %>%
  group_by(month) %>%
  summarize(mean_prct_sold = mean(prct_sold)) %>% 
  arrange(desc(mean_prct_sold))
```

Please answer the following questions with text and code. You may have to use multiple `dplyr` functions to answer each question. Think through the steps of how to get to the answer you are trying to find. First, describe the sequence of steps in words. (Perhaps you could format it as a list using markdown.) Then write the code to carry out those steps. Finally answer the question in text below the code.


9. In January of 2015, what city had the fewest houses listed for sale? (show code and text please)



10. In 2012, in which month were the most houses sold in Texas? (show code and text please)


```{r include = F}
txhousing %>% 
  filter(year == 2012) %>%
  group_by(month) %>%
  summarize(top_sales = max(sales)) %>%
  arrange(desc(top_sales))
```

11. Generate a single table that shows the total number of houses sold in **Austin** in **2000 and 2001** (total over the entire period), and the total number of houses sold in **Dallas** in **2000 and 2001** (total over the entire period). This calculation requires a number of steps, so it might help you to first write out on paper the different steps you will need to take. That will help you set out a "blueprint" for tackling the problem. **Hint**: recall the `sum()` function can add values. 


